A Classical Density-Functional Theory for Describing Water Interfaces

### 2012-07-12 papers/water-SAFT/paper.tex
Changes to paper figures (some need fixing still).  Updated two rod energy-vs-distance. Removed energy-rods.gp since it is not being used. Tested everything on node04.

A number of exact properties are easily achieved in the
density-functional framework, such as the contact-value theorem, which
ensures a correct free energy of hydration \fixme{change this to
  excess chemical potential?} for small hard solutes.
Much of the research on classical density-functional theory has
focused on the hard-sphere fluid~\cite{curtin1985, rosenfeld1989,
  rosenfeld1993, rosenfeld1997, tarazona1997, tarazona2000}, which has
led to a number of sophisticated functionals, such as the
fundamental-measure theory (FMT) functionals\cite{rosenfeld1989,
  rosenfeld1993, rosenfeld1997, tarazona1997, tarazona2000,
  roth2002whitebear}.  These functionals are entirely expressed as an integral of
local functions of a few convolutions of the density that can be
efficiently computed.  We will use the White Bear version of the FMT
functional\cite{roth2002whitebear}.  In the homogeneous limit, this
functional reduces to the Carnahan-Starling equation of state, and in
the strongly-confined limit of a highly confined cavity, it reproduces
the exact free energy.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/energy-vs-diameter}
\end{center}
\caption{ Free energy of hydration versus radius for a single hydrophobic rod
immersed in water. This should have an asymptote equal to the surface
tension at room temperature, and it agrees well with the surface tension in
Figure~\ref{fig:surface-tension}. The inset for rods with a very small 
radius shows the linear relationship expected based on 
Equation~\ref{contactvaluethm}.}
\label{fig:energy-vs-diameter}
\end{figure}

We begin by studying a single hydrophobic rod immersed in water.
% We
% model the density on a grid with spacing of about 0.01~nm, which gives
% the free energy of hydration to a precision of better than a percent,
% and the density at contact to a few percent precision.
We show in Figure~\ref{fig:energy-vs-diameter} the free energy of
hydration at room temperature, scaled by the solvent-accessible
surface area of the hard rod, plotted as a function of hard-rod
radius, which we define as the radius from which water is excluded.
For rods with radius larger than 0.5~nm or so, this reaches a value of
75~mN/m, which is slightly higher than the bulk surface, which will be
approached from above in the limit of very large rods \fixme{this is a
terrible sentence, want to say something more concrete about free
energy / area dropping (getting data) DJR}.  For very small
radius (less than about 0.1~nm) the relationship between the free
energy per area and radius is linear, with the slope required by the
contact-value theorem (see Equation~\ref{contactvaluethm}).

@@

We now look at the more interesting problem of two parallel hard rods
in water, separated by a distance $d$, as shown in
Figure~\ref{fig:rods}.  At small separations there is only vapor
between the rods, but as the rods are pulled apart, the vapor region
expands until a critical separation is reached at which point liquid
water fills the region between the rods.
Figure~\ref{fig:density-rods} shows density profiles before and after
this transition for rods of radius 0.5~nm. This critical separation
for the transition to liquid depends on the radii of the rods, and is
about 0.54~nm for the rods shown in Figure~\ref{fig:density-rods}.
The critical separation will be different for a system where there is
a significant direct force between rods, or attraction between the
rods and water.  At small separations, the water around the two
rods looks as if the two rods were one solid ``stadium''-shaped object
(a rectangle with semi-circles on both ends).

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/rods-diagram}
\end{center}
\caption{ Cross section schematic for two hydrophobic rods immersed in water.
The radius and separation are shown.}
\label{fig:rods}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/four-rods-energy-vs-distance}
\end{center}
\caption{ Free energy of hydration 
versus separation for four hydrophobic rods ranging in radius from
0.3~nm to 1.0~nm.
All were arbitrarily offset to zero at large separations. The
transition corresponds to the phase change from
vapor to liquid between the rods as pictured in the density profiles in 
Figure~\ref{fig:density-four-rods}. }
\label{fig:four-rods-energy-vs-distance}
\end{figure}

@@

We go on to study four parallel hard rods, as examined by Lum,
Chandler and Weeks in their classic paper on hydrophobicity at small
and large length scales~\cite{lum1999hydrophobicity}.  As in the case
of two rods---and as predicted by Lum~\emph{et al}---we observe a
drying transition, as seen the density plot shown in
Figure~\ref{fig:density-four-rods}.  In
Figure~\ref{fig:four-rods-energy-vs-distance}, we plot the free energy
of hydration together with the macroscopic approximation, and find
good agreement for rods larger than 1~nm in diameter.  \fixme{ Add
  data from LCW paper to the plot? I ended up cutting what felt like
  needless verbiage (and excessive explanation of the macroscopic
  model), but now it seems too short. }

@@

A common model of hydrophobic solutes is the hard-sphere solute, which
is the simplest possible solute, and serves as a test case for
understanding of hydrophobic solutes in
water\cite{sedlmeier2011entropy}.  As in the single rod, we begin by
examining the free energy of hydration, expressed in
Figure~\ref{fig:sphere-energy-vs-diameter} as the ratio of the energy
of the cavity system to the solvent-accessible surface area.  This
effective surface tension surpasses the bulk surface tension at a
radius of almost 1~nm, and will eventually drop to the bulk value.  As
with the single rod, we see the analytically correct behavior in the
limit of small solutes.  For comparison, we plot the free energy
calculated using a molecular dynamics simulation of SPC/E
water\cite{huang2001shs}.  The agreement is quite good, apart from the
issue that the SPC/E model for water significantly underestimates the
surface bulk tension of water at room
temperature\cite{vega2007surface}.

>>>>

A number of exact properties are easily achieved in the
density-functional framework, such as the contact-value theorem, which
ensures a correct excess chemical potential for small hard solutes.
Much of the research on classical density-functional theory has
focused on the hard-sphere fluid~\cite{curtin1985, rosenfeld1989,
  rosenfeld1993, rosenfeld1997, tarazona1997, tarazona2000}, which has
led to a number of sophisticated functionals, such as the
fundamental-measure theory (FMT) functionals\cite{rosenfeld1989,
  rosenfeld1993, rosenfeld1997, tarazona1997, tarazona2000,
  roth2002whitebear}.  These functionals are entirely expressed as an integral of
local functions of a few convolutions of the density that can be
efficiently computed.  We will use the White Bear version of the FMT
functional\cite{roth2002whitebear}.  In the homogeneous limit, this
functional reduces to the Carnahan-Starling equation of state, and in
the strongly-confined limit of a highly confined cavity, it reproduces
the exact free energy.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/energy-vs-diameter}
\end{center}
\caption{ \fixme{Free energy of hydration} versus radius for a single hydrophobic rod
immersed in water. This should have an asymptote equal to the surface
tension at room temperature, and it agrees well with the surface tension in
Figure~\ref{fig:surface-tension}. The inset for rods with a very small 
radius shows the linear relationship expected based on 
Equation~\ref{contactvaluethm}.}
\label{fig:energy-vs-diameter}
\end{figure}

We begin by studying a single hydrophobic rod immersed in water.
% We
% model the density on a grid with spacing of about 0.01~nm, which gives
% the free energy of hydration to a precision of better than a percent,
% and the density at contact to a few percent precision.
We show in Figure~\ref{fig:energy-vs-diameter} the excess chemical
potential \fixme{does this work?} at room temperature, scaled by the solvent-accessible
surface area of the hard rod, plotted as a function of hard-rod
radius, which we define as the radius from which water is excluded.
For rods with radius larger than 0.5~nm or so, this reaches a value of
75~mN/m, which is slightly higher than the bulk surface, which will be
approached from above in the limit of very large rods \fixme{this is a
terrible sentence, want to say something more concrete about free
energy / area dropping (getting data) DJR}.  For very small
radius (less than about 0.1~nm) the relationship between the free
energy per area and radius is linear, with the slope required by the
contact-value theorem (see Equation~\ref{contactvaluethm}).

@@

We now look at the more interesting problem of two parallel hard rods
in water, separated by a distance $d$, as shown in
Figure~\ref{fig:density-rods}.  At small separations there is only vapor
between the rods, but as the rods are pulled apart, the vapor region
expands until a critical separation is reached at which point liquid
water fills the region between the rods.
Figure~\ref{fig:density-rods} shows density profiles before and after
this transition for rods of radius 0.5~nm. This critical separation
for the transition to liquid depends on the radii of the rods, and is
about 0.54~nm for the rods shown in Figure~\ref{fig:density-rods}.
The critical separation will be different for a system where there is
a significant direct force between rods, or attraction between the
rods and water.  At small separations, the water around the two
rods looks as if the two rods were one solid ``stadium''-shaped object
(a rectangle with semi-circles on both ends).

%\begin{figure}
%\begin{center}
%\includegraphics[width=\columnwidth]{figs/rods-diagram}
%\end{center}
%\caption{ Cross section schematic for two hydrophobic rods immersed in water.
%The radius and separation are shown.}
%\label{fig:rods}
%\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/four-rods-energy-vs-distance}
\end{center}
\caption{ \fixme{Free energy of hydration} 
versus separation for four hydrophobic rods ranging in radius from
0.3~nm to 1.0~nm.
All were arbitrarily offset to zero at large separations. The
transition corresponds to the phase change from
vapor to liquid between the rods as pictured in the density profiles in 
Figure~\ref{fig:density-four-rods}. }
\label{fig:four-rods-energy-vs-distance}
\end{figure}

@@

We go on to study four parallel hard rods, as examined by Lum,
Chandler and Weeks in their classic paper on hydrophobicity at small
and large length scales~\cite{lum1999hydrophobicity}.  As in the case
of two rods---and as predicted by Lum~\emph{et al}---we observe a
drying transition, as seen the density plot shown in
Figure~\ref{fig:density-four-rods}.  In
Figure~\ref{fig:four-rods-energy-vs-distance}, we plot the \fixme{free energy
of hydration} together with the macroscopic approximation, and find
good agreement for rods larger than 1~nm in diameter.  \fixme{ Add
  data from LCW paper to the plot? I ended up cutting what felt like
  needless verbiage (and excessive explanation of the macroscopic
  model), but now it seems too short. }

@@

A common model of hydrophobic solutes is the hard-sphere solute, which
is the simplest possible solute, and serves as a test case for
understanding of hydrophobic solutes in
water\cite{sedlmeier2011entropy}.  As in the single rod, we begin by
examining the \fixme{free energy of hydration}, expressed in
Figure~\ref{fig:sphere-energy-vs-diameter} as the ratio of the energy
of the cavity system to the solvent-accessible surface area.  This
effective surface tension surpasses the bulk surface tension at a
radius of almost 1~nm, and will eventually drop to the bulk value.  As
with the single rod, we see the analytically correct behavior in the
limit of small solutes.  For comparison, we plot the free energy
calculated using a molecular dynamics simulation of SPC/E
water\cite{huang2001shs}.  The agreement is quite good, apart from the
issue that the SPC/E model for water significantly underestimates the
surface bulk tension of water at room
temperature\cite{vega2007surface}.
